---
title: "pca3D"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pca3D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(pca3D)
library(ggplot2)
```

pca3D creates 3D-style PCA scatter plots from a `prcomp` object using
ggplot2, plus rotating GIF animations of a full 360 degree spin. Because
ggplot2 has no native 3D coordinate system, pca3D simulates one: it applies
a 3D rotation to the three selected PC scores and projects the result onto
2D, optionally scaling point size and opacity by depth so closer points
stand out (a "depth cue"). A single such projection is a static plot; many
projections at different azimuth angles, stitched together, form the
rotating GIF.

## Fitting a PCA

We'll use the classic `iris` dataset, colouring points by species (a
categorical variable) and by `Sepal.Length` (a continuous variable).

```{r fit-pca}
pca <- prcomp(iris[, 1:4], scale. = TRUE)
```

## A static plot

`plot_pca3d()` returns a single view of the point cloud, at a given azimuth
(`theta`) and elevation (`phi`) angle:

```{r static-plot}
plot_pca3d(pca, metadata = iris, color_by = Species)
```

By default, the first three principal components are used, and a caption
lists the percent variance explained by each. Once the cloud is rotated,
the 2D x/y axes are a blend of the selected PCs rather than any single one,
so axis titles are intentionally omitted in favour of this caption.

### Choosing PC dimensions

Use `dims` to plot a different set of three components:

```{r dims}
plot_pca3d(pca, dims = c(1, 2, 4), metadata = iris, color_by = Species)
```

### Colouring by a continuous variable

`color_by` works the same way for numeric columns:

```{r continuous}
plot_pca3d(pca, metadata = iris, color_by = Sepal.Length)
```

### Customising the plot

`plot_pca3d()` returns a regular `ggplot` object and applies no colour
scale of its own, so you're free to add your own scale, theme, or other
layers:

```{r customise}
plot_pca3d(pca, metadata = iris, color_by = Species) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
```

### Viewing angle and depth cueing

`theta` and `phi` control the azimuth and elevation of the view, and
`depth_cue` toggles the size/opacity effect used to suggest depth:

```{r angle}
plot_pca3d(pca, metadata = iris, color_by = Species, theta = 120, phi = 10)
```

```{r no-depth-cue}
plot_pca3d(pca, metadata = iris, color_by = Species, depth_cue = FALSE)
```

### Showing which way each axis points

Once the cloud is rotated, it's easy to lose track of which direction each
PC points. Set `axes = "full"` to draw an arrow through the point cloud
along each rotated PC direction, labelled at the tip:

```{r axes-full}
plot_pca3d(pca, metadata = iris, color_by = Species, axes = "full")
```

Or `axes = "gizmo"` for a small triad of arrows tucked in a corner,
independent of the data's scale -- useful when full-length arrows would be
too cluttered, especially once the plot is animated:

```{r axes-gizmo}
plot_pca3d(pca, metadata = iris, color_by = Species, axes = "gizmo")
```

Both styles rotate together with the point cloud (in `animate_pca3d()`
too), so they always reflect the current viewing angle. `axis_color`
controls their (fixed, non-data-mapped) colour.

## Animating a full rotation

`animate_pca3d()` renders many views spanning a full 360 degree rotation
and saves them as a looping GIF. It takes largely the same arguments as
`plot_pca3d()` (minus `theta`, since every angle is rendered), plus
`n_frames`, `fps`, and output dimensions:

```{r animate, eval = FALSE}
animate_pca3d(
  pca,
  metadata = iris,
  color_by = Species,
  axes = "gizmo",
  n_frames = 36,
  fps = 10,
  file = "pca3d.gif"
)
```

## Metadata alignment

When `metadata` and the data used to fit `pca` both have meaningful row
names, pca3D matches rows by name so metadata doesn't need to be in the
same order as the original data. If either is missing row names, rows are
assumed to already be in the same order.
